---
include-in-header:
  text: |
    \newcommand{\mytitle}{High Returns, Low Investment: \\ Public Capital and the Euler Equation Puzzle}
format:
  aea-pdf:
    keep-tex: false  
    journal:
      blinded: false
    indent: true
    citecolor: red
    linkcolor: red
    urlcolor: red
  aea-html: default
author-format: horizontal
author:
  - name: "\\LARGE Yichu Li"
    acknowledgements: | 
      The London School of Economics and Political Science. [y.li395@lse.ac.uk](mailto:y.li395@lse.ac.uk).
date: last-modified
articleurl: https://tsssssssshark.github.io/euler-collective-action/output/main.pdf
abstract: |
  The text of your abstract.
keywords:
  - Intertemporal Choice, Collective Action, Public Investment
bibliography: references.bib
execute:
  echo: false
  warning: false
  message: false
  cache: false
---

```{r}
#| label: setup-packages
#| include: false

# Check and install required packages
required_packages <- c(
  "tidyverse",
  "kableExtra",
  "here",
  "glue",
  "flextable",
  "gt",
  "ggplot2",
  "haven",
  "labelled"
)

missing_packages <- required_packages[
  !required_packages %in% rownames(installed.packages())
]

if (length(missing_packages) > 0) {
  message(
    "Installing missing packages:",
    paste(missing_packages, collapse = ", "),
    "\n"
  )
  install.packages(missing_packages)
}

lapply(required_packages, library, character.only = TRUE)

message("All packages loaded successfully!")
```

## Introduction {#sec-intro}

I will show some examples of things you can easily do in quarto-format.Results

## An Euler Equation Model with Public Capital Externalities {#sec-model}

I build on the intertemporal optimisation framework of @kremer2019. I first consider the case where capital is entirely public and investment occurs only through taxation, showing that this generates a collective action problem that dilutes the effective return to capital. I then extend the standard model to incorporate both public and private capital, allowing households to allocate investments between the two.

### Baseline: Standard Euler Equation

Consider in a deterministic discrete-time intertemporal consumption model, households maximise their discounted lifetime utility over an infinite horizon.

#### Optimisation Problem {.unnumbered}

Households face the following optimisation problem:

$$
\max_{\{C_t\}} \sum_{t=0}^\infty \delta^t u(C_t)
$$ {#eq-optimisation} where $0<\delta<1$ is the annual discount factor, $C_t$ is the household consumption in period $t$, and $u(C_t)$ measures the utility from consuming $C_t$ in period $t$. $u(C)$ is increasing and concave and follows Inada Conditions: $u'(C) > 0$, $u''(C) < 0$, $\lim_{C \to 0} u'(C) = \infty$, and $\lim_{C \to \infty} u'(C) = 0$.

#### Budget Constraint {.unnumbered}

$$
x_{t+1} = f'(K_t)(x_t - C_t)
$$ {#eq-budget}$x_t$ is household assets ('cash on hand') in period $t$ with $x_0$ given, $C_t$ is consumption in period $t$, and $f'(K_t)$ is the gross return to capital. $f(K)$ is increasing and concave, following: $f'(K) > 0$, $f''(K) < 0$, $\lim_{K \to 0} f'(K) = \infty$, and $\lim_{K \to \infty} f'(K) = 1$ (long-run steady state).

#### Discrete Time Euler Equation {.unnumbered}

Solving for optimal intertemporal consumption path:

$$
u'(C_t) = \delta f'(K_t) u'(C_{t+1})
$$ {#eq-euler} Assume constant-elasticity-of-substitution (CES) utility:

$$
u(C) = 
\begin{cases}
    \frac{C^{1-\sigma}}{1-\sigma} & \text{if } \sigma \ne 1\\
    \ln(C) & \text{if } \sigma = 1
\end{cases}
$$This means that the elasticity of intertemporal substitution is constant and independent of $C_t$. Therefore, the consumption growth rate is:

$$
\frac{C_{t+1}}{C_t} - 1 = (\delta f'(K_t))^{\frac{1}{\sigma}} - 1
$$ {#eq-growth}

#### The Puzzle in Low Income Countries {.unnumbered}

Using $\delta = 0.96$ \footnote{$\frac{1}{\delta} - 1 \approx 0.04$, implying a 4\% annual rate of time preference.}and $\sigma = 2$ \footnote{Elasticity of intertemporal substitution of 0.5 is used in macro studies standard calibrations.} proposed in @kremer2019 as an example, and predicting $f'(K)$ using real internal rate of return, @fig-deviation shows that countries with higher predicted consumption growth rates tend to have actual consumption growth rates below the predicted rates.

```{r}
#| label: fig-deviation
#| eval: true
#| fig-cap: "Predicted vs. Actual Consumption Growth"

# Read PWT dataset
pwt <- read_dta(here("data", "pwt110", "pwt110.dta"))

# Predicted vs. real consumption growth rate using only 'regular' irr
pwt_dev <- pwt %>%
  filter(i_irr == 0) %>%
  mutate(
    f_prime_k = 1 + irr,
    g_predicted = (0.96 * f_prime_k)^(1/2) - 1,
    cons_pc = rconna / pop
  ) %>%
  # Average
  group_by(countrycode) %>%
  arrange(year) %>%
  mutate(g_actual = (cons_pc / lag(cons_pc)) - 1) %>%
    summarise(
      g_predicted = mean(g_predicted, na.rm = TRUE),
      g_actual = mean(g_actual, na.rm = TRUE),
      .groups = "drop"
    )

# Graph
ggplot(pwt_dev, aes(x = g_predicted, y = g_actual)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "Predicted Consumption Growth Rate",
    y = "Actual Consumption Growth Rate"
  ) +
  theme_classic() +
  theme(
    plot.margin = margin(t = 5, r = 15, b = 5, l = 5),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5),
    axis.ticks.length = unit(-0.15, "cm"),
    axis.text.x = element_text(margin = margin(t = 0.3, unit = "cm")),
    axis.text.y = element_text(margin = margin(r = 0.3, unit = "cm")),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )
```

## Conclusion

\newpage

## References {.unnumbered}
