---
include-in-header:
  text: |
    \newcommand{\mytitle}{Public Capital Investment Frictions: \\ A Piece in the Euler Equation Puzzle}
    \usepackage{parskip}
    \setlength{\parskip}{0.5\baselineskip}
    \setlength{\parindent}{2em}
    \usepackage{pdflscape}
    \usepackage{rotating}
format:
  aea-pdf:
    keep-tex: false  
    journal:
      blinded: false
    indent: true
    citecolor: red
    linkcolor: red
    urlcolor: red
    reference-location: section
  aea-html: default
author-format: horizontal
author:
  - name: "\\LARGE Yichu Li"
    acknowledgements: | 
      The London School of Economics and Political Science. [y.li395@lse.ac.uk](mailto:y.li395@lse.ac.uk).
date: last-modified
articleurl: https://tsssssssshark.github.io/euler-collective-action/output/main.pdf
abstract: |
  The text of your abstract.
keywords:
  - Growth, Euler Equation Puzzle, Public Investment, Development Accounting
bibliography: references.bib
execute:
  echo: false
  warning: false
  message: false
  cache: false
---

```{r}
#| label: setup-packages
#| include: false

# Check and install required packages
required_packages <- c(
  "tidyverse",
  "kableExtra",
  "here",
  "glue",
  "flextable",
  "gt",
  "ggplot2",
  "haven",
  "labelled",
  "patchwork",
  "countrycode",
  "fixest",
  "modelsummary",
  "diptest",
  "plm"
)

missing_packages <- required_packages[
  !required_packages %in% rownames(installed.packages())
]

if (length(missing_packages) > 0) {
  message(
    "Installing missing packages:",
    paste(missing_packages, collapse = ", "),
    "\n"
  )
  install.packages(missing_packages)
}

lapply(required_packages, library, character.only = TRUE)

message("All packages loaded successfully!")
```

## Introduction {#sec-intro}

Much of development economics research attempt to understand what is holding back economic development in lower- and middle-income countries. The categorisation of countries into 'developing' versus 'developed' implies that the former have higher development potentials. The @solow1956 growth model predicts exactly this pattern because poorer economies with smaller capital stocks should face higher marginal returns to capital investment and therefore grow faster than richer economies. While the broad pattern of higher GDP growth rates in lower-income countries over the past four decades is consistent with this prediction, many countries have still believed to been growing at rates below their potentials.

@lucas1990 and @caselli2007 points out that the higher marginal product of capital in lower-income countries is often balanced-off by their lower endowment in complimentary factors like human capital, or lower total factor productivity. This provides an aggregate-level explanation for why capitals do not flow from higher-income to lower-income countries as much as expected. However, the puzzle is not fully addressed because on the aggregate-level, it does not explain why lower-income countries do not invest in human capital or R&D to benefit from the high capital returns. Moreover, through micro-level RCTs, @demel2008 show that microenterprises in Sri Lanka face 55%-63% real anual returns to capital, and @balboni2022 show significant long-run returns on agricultural capitals sufficient to push households in Bangladesh out of a poverty trap. @kremer2019 uses an household-level intertemporal constrained optimisation model to illustrate an '*Euler Equation Puzzle*' where people in lower-income settings should experience a 20% consumption growth rate facing a 150% gross return, but the actual consumption growth rates are much lower.

Relating to this paradox, a large body of literature including @balboni2022 and research covered in @banerjee2003 point to liquidity and credit constraints facing households and firms; some including @acemoglu2001, @mauro1995 and @banerjee2002 point to institutional ineffectiveness and insufficient property rights protections; while @kremer2019 point to studies in behavioural channels such as present bias. However, the existing literature focus primarily on private investment channels. For example, these are scenarios where farmers investing in a cattle or fertilisers and benefit fully from the output gains. Meanwhile, returns on certain capitals like infrastructure or human capital are also high in lower-income countries. Investment in public capitals often require fiscal spending from the government funded primarily by tax revenues. @besley2013 show that in lower-income countries taxation is more difficult administratively while frictions against efficient spending are stronger. The constraints on both private and public capital investment seem to be likely correlated with the existing capital accumulation and development levels, hinting towards an endogenous capital investment mechanism that relate to the poverty trap and convergence literature. Studies like @balboni2022 shows that poverty traps likely bind the ultra-poor households to invest, while research on aggregate growth like @quah1996 documents the bimodal pattern and @aghion2005 identifies the role of financial development in threshold convergence.

Nonetheless, @kremer2019 questions why households would not rationally attempt more to smooth consumption or leverage on informal credits when motivated by the high returns. The poverty traps thresholds are likely less binding for the non-ultra-poor, but they systematically not exploit the capital returns sufficiently either. The same question extends to the public capital investment: it seems irrational for above-subsistence households in lower-income countries to avoid taxes even with low expected penalties because the long-term returns from public capitals can be sufficiently high. This paper adapts and develop on the baseline mode of @kremer2019 to show that the set-up of the taxation-to-government-spending mechanism produces a collective action problem at the household level. Public capital investment requires highly lumpy fiscal spending funded by taxation, and the total returns are shared by each household like public goods. When the perceived state taxation capacity is low, households expect low actual returns on public capital investment via taxation because they cannot credibly expect others to pay taxes symmetrically. Therefore, for sufficiently low state taxation capacity, households' dominant strategy is to consume all and pay zero taxes.

## A Household Intertemporal Constrained Optimisation Model on Capital Investment {#sec-model}

I build on the intertemporal optimisation framework of @kremer2019. I first recall the model setup and present the Euler Equation Puzzle through the model. I then consider the extreme scenario where only public capitals exist and investment occurs only through taxation, showing that this generates a collective action problem and a government efficiency problem that dilutes the effective return to capital when state taxation capacity and efficiency is low. Finally, I will extend this analysis into a generalised model incorporating both private and public capital, allowing households to allocate investments between the two.

### Baseline: Standard Euler Equation {#sec-baseline}

Consider in a deterministic discrete-time intertemporal consumption model, households maximise their discounted lifetime utility over an infinite time horizon.

#### Optimisation Problem {.unnumbered}

Households face the following optimisation problem: $$\max_{\{C_t\}} \sum_{t=0}^\infty \delta^t u(C_t),$$ {#eq-optimisation} where $0<\delta<1$ is the annual discount factor, $C_t$ is the household consumption in period $t$, and $u(C_t)$ measures the utility from consuming $C_t$ in period $t$. $u(C)$ is increasing and concave and follows Inada Conditions: $u'(C) > 0$, $u''(C) < 0$, $\lim_{C \to 0} u'(C) = \infty$, and $\lim_{C \to \infty} u'(C) = 0$.

#### Budget Constraint {.unnumbered}

While the model set-up in @kremer2019 consider the gross return to capital, I consider the return to capital which allows me to model capital depreciation that the PWT data provides at the aggregate level. Therefore, the budget constraint is:

$$x_{t+1} = [(1-d) + f'(K_t)](x_t - C_t),$$ {#eq-budget} $x_t$ is household assets ('cash on hand') in period $t$ with $x_0$ given, $C_t$ is consumption in period $t$, and $f'(K_t)$ is the return to capital, and $d$ is the capital depreciation rate. $f(K)$ is increasing and sconcave, following: $f'(K) > 0$, $f''(K) < 0$, $\lim_{K \to 0} f'(K) = \infty$, and $\lim_{K \to \infty} f'(K) = 0$ (long-run steady state).

#### Discrete Time Euler Equation {.unnumbered}

Solve for optimal intertemporal consumption path: $$u'(C_t) = \delta \cdot [(1-d) + f'(K_t)] \cdot u'(C_{t+1}),$$ {#eq-euler} and assume constant-elasticity-of-substitution (CES) utility: $$u(C) = 
\begin{cases}
    \frac{C^{1-\sigma}}{1-\sigma} & \text{if } \sigma \ne 1\\
    \ln(C) & \text{if } \sigma = 1
\end{cases}$$ so that the elasticity of intertemporal substitution is constant and independent of $C_t$. This gives the consumption growth rate: $$\frac{C_{t+1}}{C_t} - 1 = (\delta [(1-d) + f'(K_t)])^{\frac{1}{\sigma}} - 1,$$ {#eq-growth} where $\frac{1}{\sigma}$ is the elasticity of intertemporal substitution.

#### The Puzzle {.unnumbered}

Assume that the model governs the intertemporal consumption behaviour of a representative household, the model should also apply at the macro level. To illustrate the puzzle with macro-level data, I use $\delta = 0.96$ \footnote{$\frac{1}{\delta} - 1 \approx 0.04$, implying a 4\% annual rate of time preference.}and $\sigma = 2$ \footnote{Elasticity of intertemporal substitution of 0.5.} proposed in @kremer2019, and internal rates of return (IRR) from PWT \footnote{Available for download at \url{www.ggdc.net/pwt}} [@feenstra2015a] as a proxy for $f'(K)$\footnote{See \url{https://www.rug.nl/ggdc/docs/the_next_generation_of_the_penn_world_table.pdf}} to model the predicted consumption growth rates under @eq-growth. Intuitively, a concave production function means that countries with lower capital stock face higher returns on capital investment. @fig-k-gdp shows that countries with lower income have lower capital stock which is expected because the @solow1956 model shows that output increases with capital accumulation. Therefore, according to @eq-growth, consumption growth rates should be high in lower-income countries. However, @fig-actual-predict Panel (A) shows that in many countries, the actual growth rates is insensitive to IRR and many fall below the model predictions. Moreover, @fig-actual-predict Panel (B) shows a perplex trend where countries with higher predicted consumption growth rates tend to have larger gap between predicted and actual rates. This pattern is most pronounced in countries lower-middle and low income countries. Therefore, the core of the puzzle are why households in these economies invest at levels far below what the standard model predicts as optimal, and why is this pattern more pronounced in countries where returns on capitals are high.

```{r}
#| label: read-pwt

# Read PWT dataset
pwt <- read_dta(here("data", "pwt110", "pwt110.dta")) %>%
  mutate(
    gdp_pc = rgdpo / pop,
    capital_pc = cn / pop
  )

small_countries <- pwt %>%
  filter(year == 2019) %>%
  filter(pop < 1) %>% 
  pull(countrycode)
```

```{r}
#| label: fig-k-gdp
#| fig-cap: "Lower Capital Stock in Lower-Income Countries"
#| fig-cap-location: top

pwt %>%
  filter(!is.na(rgdpe), !is.na(rkna), !is.na(pop), pop >= 1) %>%
  ggplot(aes(x = log(gdp_pc), y = log(capital_pc))) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(
    x = "Log GDP per Capita (PPP)",
    y = "Log Capital Stock per Capita",
    caption = stringr::str_wrap("Notes: Red line is OLS fit. Panel data from 1950 to 2023. Capital stock in millions US$ at 2021 PPPs. Output-side real GDP in millions US$ at 2021 chained PPPs. Source: PWT 11.0.", width = 100)
  ) +
  theme_classic() +
  theme(
    plot.caption = element_text(hjust = 0, size = 8, face = "italic"),
    plot.caption.position = "plot",
    plot.margin = margin(t = 5, r = 15, b = 5, l = 5),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5),
    axis.ticks.length = unit(-0.15, "cm"),
    axis.text.x = element_text(margin = margin(t = 0.3, unit = "cm")),
    axis.text.y = element_text(margin = margin(r = 0.3, unit = "cm"))
  )
```

```{r}
#| label: fig-actual-predict
#| eval: true
#| fig-cap: "Actual vs. Predicted Consumption Growth"
#| fig-cap-location: top
#| fig-width: 10
#| fig-height: 5

pwt_dev <- pwt %>%
  filter(i_irr == 0, year <= 2019, !countrycode %in% small_countries) %>%
  group_by(countrycode) %>%
  arrange(year) %>%
  mutate(
    f_prime_k = dplyr::lag(irr),
    g_predicted = (0.96 * ((1-delta) + f_prime_k))^(1/2) - 1,
    cons_pc = rconna / pop,
    g_actual = (cons_pc / dplyr::lag(cons_pc)) - 1
  ) %>%
  ungroup() %>%
  filter(!is.na(g_predicted), !is.na(g_actual)) %>%
  mutate(
    log_g_predicted = log(1 + g_predicted),
    log_g_actual = log(1 + g_actual),
    gap = log_g_predicted - log_g_actual
  )

make_binscatter <- function(data, x_var, y_var, n_bins = 100) {
  data %>%
    mutate(bin = ntile(.data[[x_var]], n_bins)) %>%
    group_by(bin) %>%
    summarise(
      x = mean(.data[[x_var]], na.rm = TRUE),
      y = mean(.data[[y_var]], na.rm = TRUE)
    )
}

bin_data_a1 <- make_binscatter(pwt_dev, "log_g_predicted", "log_g_actual")
axis_range <- range(c(bin_data_a1$x, bin_data_a1$y))

a1 <- bin_data_a1 %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "blue") +
  xlim(axis_range) +
  ylim(axis_range) +
  coord_fixed(ratio = 1) +
  labs(
    x = "Log Gross Predicted Consumption Growth Rate",
    y = "Log Gross Actual Consumption Growth Rate",
    title = "(A)",
    caption = stringr::str_wrap("Notes: Binned scatter plots with 100 equal-frequency bins. The blue dashed line is the 45 degree line. The red line is OLS fit. Countries with populations below 1 million and irregular internal rates of return excluded. Data after 2019 is excluded to avoid Covid-19 distortions. Source: PWT 11.0.", width = 190)
  ) +
  theme_classic() +
  theme(
    plot.margin = margin(t = 5, r = 15, b = 5, l = 5),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5),
    axis.ticks.length = unit(-0.15, "cm"),
    axis.text.x = element_text(margin = margin(t = 0.3, unit = "cm")),
    axis.text.y = element_text(margin = margin(r = 0.3, unit = "cm")),
    plot.caption = element_text(hjust = 0, size = 8, face = "italic"),
    plot.caption.position = "plot",
    plot.title = element_text(hjust = 0.5)
  )

b1 <- make_binscatter(pwt_dev, "log_g_predicted", "gap") %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  coord_fixed(ratio = 1) +
  labs(
    x = "Log Gross Predicted Consumption Growth Rate",
    y = "Log Gross Growth Rate Gap",
    title = "(B)"
  ) +
  theme_classic() +
  theme(
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5),
    axis.ticks.length = unit(-0.15, "cm"),
    axis.text.x = element_text(margin = margin(t = 0.3, unit = "cm")),
    axis.text.y = element_text(margin = margin(r = 0.3, unit = "cm")),
    plot.caption = element_text(hjust = 0, size = 8, face = "italic"),
    plot.caption.position = "plot",
    plot.title = element_text(hjust = 0.5)
  )

a1 + b1 
```

### Endogenous Capital Accumulation {#sec-historical-accumulation}

There is a potential concern for the findings in @fig-actual-predict that the insensitivity of the actual consumption growth rate towards the model prediction signals severe model mis-specification. This would mean that the Euler equation model setup in @sec-baseline cannot capture the true underlying mechanism of consumption growth from capital investment, at least under aggregate data. However, the orthogonal polynomial regressions in @tbl-poly-fit show significant quadratic correlation between the actual and predicted log gross consumption growth rate, while the linear term correlation is not statistically significant . This suggests that a non-trivial relationship between predicted and actual growth does exist. Nonetheless, the small $R^2$ across specifications shows that the baseline model explains only a tiny fraction of the variation in actual consumption growth.\footnote{Appendix Table A1 reports significantly higher $R^2$ after adding per captia GDP growth and TFP growth as controls, suggesting more variations are explained by these factors.} Therefore, understanding the puzzle will require identifying the factors that systematically prevent households from exploiting high capital investment returns according to the baseline model prediction.

```{r}
#| label: tbl-poly-fit
#| tbl-cap: "Polynomial Fit of Actual on Predicted Consumption Growth"

reg_data <- pwt_dev %>%
  filter(!is.na(log_g_predicted), !is.na(log_g_actual))

poly_models <- list(
  "Linear" = lm(log_g_actual ~ log_g_predicted, data = reg_data),
  "Quadratic" = lm(log_g_actual ~ poly(log_g_predicted, 2), data = reg_data),
  "Cubic" = lm(log_g_actual ~ poly(log_g_predicted, 3), data = reg_data)
)

coef_names <- c(
  "log_g_predicted" = "Log Predicted Growth",
  "poly(log_g_predicted, 2)1" = "Poly term 1",
  "poly(log_g_predicted, 2)2" = "Poly term 2",
  "poly(log_g_predicted, 3)1" = "Poly term 1",
  "poly(log_g_predicted, 3)2" = "Poly term 2",
  "poly(log_g_predicted, 3)3" = "Poly term 3",
  "(Intercept)" = "Intercept"
)

stars <- c('*' = .05, '**' = .01, '***' = .001)
gof <- data.frame(
  raw = c("nobs", "r.squared"),
  clean = c("Sample Size", "$R^2$"),
  fmt = c("%.0f", "%.4f")
)

modelsummary(
  poly_models,
  coef_map = coef_names,
  stars = stars,
  gof_map = gof,
  output = "kableExtra",
  escape = FALSE,
  align = "lccc"
) %>%
  add_header_above(c(" " = 1, "Log Actual Consumption Growth" = 3)) %>%
  kable_styling(latex_options = "HOLD_position") %>%
  gsub("\\\\end\\{tabular\\}",
     "\\\\end{tabular}\n\\\\begin{minipage}{\\\\textwidth}\n\\\\vspace{2pt}\n\\\\footnotesize \\\\textit{Notes:} Orthogonal polynomial regressions of log actual on log predicted gross consumption growth. Standard errors in parentheses. \n\\\\end{minipage}", .)
```

A mathematically plausible way to understand the reason for @fig-actual-predict Panel (B) is to consider that households or countries that tend to under-invest in capital likely simultaneously have lower consumption growth and lower capital stocks. This means that the capital return rates should be high due to low capital stocks which predict high consumption growth, but the actual consumption growth is low because they invest sub-optimally, therefore the consumption growth gap is large. The countries with higher predicted consumption growth in @fig-actual-predict Panel (B) tend to be countries that under-invest relative to the predicted levels, have lower capital stocks, and exhibit larger consumption growth gaps. @tbl-gap-poly-fit also shows that the linear pattern is predominantly driven by large linear term coefficient while the quadratic coefficient is statistically significant but too small to be visible in the binned scatter plot.

```{r}
#| label: tbl-gap-poly-fit
#| tbl-cap: "Polynomial Fit of Gap on Predicted Consumption Growth"

reg_data <- pwt_dev %>%
  filter(!is.na(gap), !is.na(log_g_actual))

poly_models <- list(
  "Linear" = lm(gap ~ log_g_predicted, data = reg_data),
  "Quadratic" = lm(gap ~ poly(log_g_predicted, 2), data = reg_data),
  "Cubic" = lm(gap ~ poly(log_g_predicted, 3), data = reg_data)
)

coef_names <- c(
  "log_g_predicted" = "Log Predicted Growth",
  "poly(log_g_predicted, 2)1" = "Poly term 1",
  "poly(log_g_predicted, 2)2" = "Poly term 2",
  "poly(log_g_predicted, 3)1" = "Poly term 1",
  "poly(log_g_predicted, 3)2" = "Poly term 2",
  "poly(log_g_predicted, 3)3" = "Poly term 3",
  "(Intercept)" = "Intercept"
)

stars <- c('*' = .05, '**' = .01, '***' = .001)
gof <- data.frame(
  raw = c("nobs", "r.squared"),
  clean = c("Sample Size", "$R^2$"),
  fmt = c("%.0f", "%.4f")
)

modelsummary(
  poly_models,
  coef_map = coef_names,
  stars = stars,
  gof_map = gof,
  output = "kableExtra",
  escape = FALSE,
  align = "lccc"
) %>%
  add_header_above(c(" " = 1, "Log Actual Consumption Growth" = 3)) %>%
  kable_styling(latex_options = "HOLD_position") %>%
  gsub("\\\\end\\{tabular\\}",
     "\\\\end{tabular}\n\\\\begin{minipage}{\\\\textwidth}\n\\\\vspace{2pt}\n\\\\footnotesize \\\\textit{Notes:} Orthogonal polynomial regressions of log gap (actual - predicted) on log predicted consumption growth. Standard errors in parentheses. \n\\\\end{minipage}", .)
```

While this intuition still does not resolve the puzzle in @fig-actual-predict Panel (A) of why some countries systematically under-invest given high returns, it provides an important insight for the likely endogeneity in the capital accumulation process that is consistent with the quadratic correlation in @tbl-poly-fit. Existing literature has studied extensively sources of frictions such as credit market or institutional inefficiencies that hinder optimal capital investment given its returns. The intuition above suggests that these frictions are likely endogenous to the capital stock level, where more frictions exist in low-capital countries and the frictions cause them to accumulate capital slowly and persist with low capital stocks. Moreover, the quadratic correlation form in @tbl-poly-fit indicates that the countries on the lower and higher ends of predicted consumption growth rate experience lower actual growth, while the countries in the medium range experience the higher growth.

### Capital as a Public Good {#sec-capital-public}

The baseline model in @sec-baseline assumes capital investments are private for households. This means that households receive the full returns $f'(K)$ because they would own the invested capital in productions. However, capitals with public good features like infrastructure often instead require public investment commonly through government spending, and the benefits from these investments are shared among the population. @aschauer1989 estimates a high average elasticity of 0.39 between output and public investment in the United States over 1949-1985, while @bom2014 proposes a more modest estimate of 0.106. The latter estimate that the short-run national-level elasticity of 0.083 still imply a 16% marginal return to public capital in OECD countries. The returns are likely higher in regional levels and much higher in lower-income countries with lower public capital stocks. However, @besley2013 show likely political economy reasons including political instability, favourtism, inequality, or corruptions which undermine the taxation-to-government-spending mechanism necessarily for optimal public capital investment. These factors are more prominent in lower-income countries, while countries might improve taxation capacities and have reduced spending frictions over development.

However, if public capital investments yield high returns, why would rational households even not pay taxes voluntarily but choose to avoid them when possible? Indeed, findings in @mauro1995 also show low public (and private) investment levels when governments are inefficient, and @fig-gove-gdp Panel (A) shows that government effectiveness is lower in lower-income countries. Therefore, this high government spending friction could discourage households to pay taxes if they believe the revenue is not put in use efficiently. Nonetheless, @fig-gove-gdp Panel (B) shows that the trend in @fig-actual-predict Panel (A) persist even controlling for government effectiveness. My following model will present that the household level, the fiscal features of lower-income countries can also create a collective action problem so that they optimally avoid paying taxes individually, thereby producing a global sub-optimal outcome.

```{r}
#| label: data-gov-e

gov_e <- read_csv(
  here("data", "wdi", "API_GE", "API_GE.EST_DS2_en_csv_v2_4300.csv"),
  skip = 3, show_col_types = FALSE
  ) %>%
  select(`Country Code`, where(is.numeric)) %>%
  pivot_longer(
    cols = -`Country Code`,
    names_to = "year",
    values_to = "gov_e"
  ) %>%
  mutate(year = as.numeric(year)) %>%
  rename(iso3c = `Country Code`)

gov_e_gdp <- gov_e %>%
  left_join(pwt %>% select(countrycode, year, rgdpo, pop),
            by = c("iso3c" = "countrycode", "year")) %>%
  mutate(gdp_pc = rgdpo / pop) %>%
  drop_na() %>%
  filter(!iso3c %in% small_countries)

resid_pwt <- pwt_dev %>%
  left_join(gov_e, by = c("countrycode" = "iso3c", "year")) %>%
  filter(!is.na(gov_e), !is.na(log_g_predicted), !is.na(log_g_actual)) %>%
  mutate(
    resid_predicted = resid(lm(log_g_predicted ~ gov_e)),
    resid_actual = resid(lm(log_g_actual ~ gov_e)),
    resid_gap = resid_predicted - resid_actual
  )
```

```{r}
#| label: fig-gove-gdp
#| fig-cap: "Gap Persists after Controlling for Government Effectiveness"
#| fig-cap-location: top
#| fig-width: 10
#| fig-height: 5

gove1 <- gov_e_gdp %>% ggplot(aes(x = log(gdp_pc), y = gov_e)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(
    x = "Log GDP per Capita (PPP)",
    y = "Government Effectiveness",
    title = "(A)",
    caption = stringr::str_wrap("Notes: Red line is OLS fit. Panel data from 1996 to 2023. Output-side real GDP in millions US$ at 2021 chained PPPs. Panel B shows binned scatter plots with 100 equal-frequency bins after partialling out government effectiveness. The blue dashed line is the 45 degree line. Source: PWT 11.0, WGI. Source: WDI.", width = 190)
  ) +
  theme_classic() +
  theme(
    plot.caption = element_text(hjust = 0, size = 8, face = "italic"),
    plot.caption.position = "plot",
    plot.title = element_text(hjust = 0.5),
    plot.margin = margin(t = 5, r = 15, b = 5, l = 5),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5),
    axis.ticks.length = unit(-0.15, "cm"),
    axis.text.x = element_text(margin = margin(t = 0.3, unit = "cm")),
    axis.text.y = element_text(margin = margin(r = 0.3, unit = "cm"))
  )

bin_data_gove2 <- make_binscatter(resid_pwt, "resid_predicted", "resid_actual")
axis_range <- range(c(bin_data_gove2$x, bin_data_gove2$y))

gove2 <- bin_data_gove2 %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "blue") +
  xlim(axis_range) +
  ylim(axis_range) +
  coord_fixed(ratio = 1) +
  labs(
    x = "Log Gross Predicted Consumption Growth Rate",
    y = "Log Gross Actual Consumption Growth Rate",
    title = "(B)"
  ) +
  theme_classic() +
  theme(
    plot.margin = margin(t = 5, r = 15, b = 5, l = 5),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5),
    axis.ticks.length = unit(-0.15, "cm"),
    axis.text.x = element_text(margin = margin(t = 0.3, unit = "cm")),
    axis.text.y = element_text(margin = margin(r = 0.3, unit = "cm")),
    plot.caption = element_text(hjust = 0, size = 8, face = "italic"),
    plot.caption.position = "plot",
    plot.title = element_text(hjust = 0.5)
  )

gove1 + gove2
```

To illustrate, I will present an extreme case where we assume capitals are all publicly invested through taxation by governments. Consider the same optimisation problem as @eq-optimisation with a different budget constraint for homogeneous households in an economy. For any household $i$, the budget constraint is: $$x_{i,t+1} = \frac{1}{N} \cdot [(1-d) + f'(K_t)] \cdot [(x_{i,t} - C_{i,t}) + \sum_{\forall j \neq i} (x_{j,t} - C_{j, t})],$$ {#eq-budget-public} where $N$ is the total number of households in the economy, and $\sum_{\forall j \neq i} (x_{j,t} - C_{j, t})$ is the aggregate public investment from all other households. Because public capitals can only be invested collectively, the total returns depend on the aggregate investment level. For simplicity, assume each individual household receive an equal share of the total returns, public capital investment is the only government expenditure, and the government has perfect efficiency.

#### Weak Taxation Capacity {.unnumbered}

First consider a scenario where the state taxation capacity is weak. To reflect weak taxation capacity, assume that household $i$ makes public investment decision through taxation voluntarily and independent of other households' decisions. From the perspective of household $i$, the total returns from public capital investment as a public good is diluted among the population. Alternatively, for a fixed level of total returns on public capital, the impact of the household's investment level $(x_{i,t} - C_{i,t})$ on its assets in the next period is small for large $N$.

Solving for optimal intertemporal consumption path for household $i$ given $\sum_{\forall j\ne i} (x_{jt} - C_{jt})$: $$u'(C_{i,t}) = \frac{1}{N} \cdot \delta \cdot [(1-d) + f'(K_t)] \cdot u'(C_{i, t+1}).$$ {#eq-euler-public} Notice that when $N=1$, this equation is identical to @eq-euler, but $u'(C_{i.t})$ decreases as $N$ becomes larger. For increasing and concave utility function, $u'(C_{i,t})$ is smaller when $C_{i,t}$ is larger. This means that household $i$ will consume more and invest less as $N$ increases. For sufficiently large $N$ so that the Euler Equation cannot be satisfied by an interior solution, household $i$ will optimally choose a corner solution of consuming all and invest none. $\frac{1}{N}$ produces a collective action problem where the global optimal decision is for all households to invest and benefit from high $f'(K)$ but individuals households have a dominant strategy of not investing.

@fig-tax-gdp shows the countries with lower GDP per capita have lower tax revenue relative to their GDP. Therefore, this scenario is more consistent with lower-income settings and it is consistent with the reluctant tax compliance attitude among many low- and middle-income settings. This framework also provides an explanation to the prima facie irrationality of tax avoidance in lower-income countries where people can benefit largely from public investments.

```{r}
#| label: data-tax

tax <- read_csv(
  here("data", "wdi", "API_GC", "API_GC.TAX.TOTL.GD.ZS_DS2_en_csv_v2_962.csv"),
  skip = 3, show_col_types = FALSE
  ) %>%
  select(`Country Code`, where(is.numeric)) %>%
  pivot_longer(
    cols = -`Country Code`,
    names_to = "year",
    values_to = "tax_per_gdp"
  ) %>%
  mutate(year = as.numeric(year)) %>%
  rename(iso3c = `Country Code`) %>%
  # Filter extreme values
  filter(tax_per_gdp > 0, tax_per_gdp < 60)

tax_gdp <- tax %>%
  left_join(pwt %>% select(countrycode, year, rgdpo, pop),
            by = c("iso3c" = "countrycode", "year")) %>%
  mutate(gdp_pc = rgdpo / pop) %>%
  drop_na() %>%
  filter(!iso3c %in% small_countries)
```

```{r}
#| label: fig-tax-gdp
#| fig-cap: "Higher Tax Revene in Higher-Income Countries"
#| fig-cap-location: top

tax_gdp %>%
  ggplot(aes(x = log(gdp_pc), y = tax_per_gdp)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(
    x = "Log GDP per Capita (PPP)",
    y = "Tax Revenue (% of GDP)",
    caption = stringr::str_wrap("Notes: Red line is OLS fit. Panel data from 1972 to 2023. Output-side real GDP in millions US$ at 2021 chained PPPs. Source: WDI.", width = 100)
  ) +
  theme_classic() +
  theme(
    plot.caption = element_text(hjust = 0, size = 8, face = "italic"),
    plot.caption.position = "plot",
    plot.margin = margin(t = 5, r = 15, b = 5, l = 5),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5),
    axis.ticks.length = unit(-0.15, "cm"),
    axis.text.x = element_text(margin = margin(t = 0.3, unit = "cm")),
    axis.text.y = element_text(margin = margin(r = 0.3, unit = "cm")),
    plot.title = element_text(hjust = 0.5)
    )
```

#### Strong Taxation Capacity {.unnumbered}

Higher income countries tend to have stronger taxation capacity because, for example, income tax can be more easily collected from employees [@jensen2022]. Although, public investment is less voluntary for tax rates as given, households might make public investment decisions by influencing the tax rates through political channels. The key difference is that they are more likely to expect other households to also comply with the decided upon tax rates.

Consider a simple case of homogeneous households facing the same optimisation problem as @eq-optimisation but assume instead perfect enforcability upon any chosen taxation level and perfect government efficiency. Any individual household $i$ chooses a taxation level (public investment level) $(x_{i,t} - C_{it})$ and expect symmetry from other households. Because households are homogeneous, the following should hold: $$[(x_{i,t} - C_{i,t}) + \sum_{\forall j \neq i} (x_{j,t} - C_{j, t})] = N \cdot (x_{i,t} - C_{i,t}),$$ so that the budget constraint becomes identical to @eq-budget: $$x_{i,t+1} = \frac{1}{N} \cdot [(1-d) + f'(K_t)] \cdot N \cdot (x_{i,t} - C_{i,t}) = [(1-d) + f'(K_t)](x_{i,t} - C_{i,t}).$$ Solving for optimal consumption path will produce an Euler Equation identical to @eq-euler. Notice that the collective action problem is resolved in this setting because the perfect tax enforcement enables households to optimise consumption conditional on the expectation of symmetrical behaviours from others.

### A Generalised Static Model {#sec-generalised-model}

The collective action problem explored above is an example channel to explain why public capital investment frictions exist. In reality, there are likely many interacted channels and the specific functional form is difficult to define. Even for the collective action model, it is very possible that the severity of the problem is not linearly sensitive to the population size in a country. Nonetheless, the public capital investment frictions at the household-decision level can be generalised as a wedge $\phi_{pub} \in [0, 1]$ where lower values indicate larger frictions. Moreover, recall that the literature also highlights various channels for the private-side frictions, so that the private-side frictions can be also generalised by $\phi_{priv} \in [0, 1]$ analogously. A generalised intertemporal household opitmisation model with both private and public capital returns has the following budget constraint: $$x_{t+1} = [(1-d) +r_t^{eff}] \cdot (x_t - C_t),$$ {#eq-budget-combined} where $$r_t^{eff} = (1-\tau) \cdot \phi^{priv} \cdot f'(K^{priv}_t) + \tau \cdot \phi^{pub} \cdot f'(K^{pub}_t),$$ with $K^{priv}$ denoting private capital stock and $\tau$ denoting the share of the total investment amount that the household chooses to invest in public capital. Notice that households choose $\tau$ freely to maximise their budget constraint. Because $f'(K_t^{priv})$ and $f'(K_t^{pub})$ are concave, this means that at equilibrium: $$\phi^{priv} \cdot f'(K_t^{priv}) = \phi^{pub} \cdot f'(K_t^{pub})$$ and rearranging: $$\frac{\phi^{pub}}{\phi^{priv}} = \frac{f'(K_t^{priv})}{f'(K_t^{pub})}$$ {#eq-ff} shows that the ratio between the public and private capital investment frictions can be given by the ratio between the marginal investment returns from private and public capital.

Solving for optimal consumption path under the above constraint produces an Euler Equation: $$u'(C_t) = \delta \cdot [(1-d) +r_t^{eff}] \cdot u'(C_{t+1}),$$ and the consumption growth rate is given by:

$$\frac{C_{t+1}}{C_t} - 1 = \{\delta [(1-d) +r_t^{eff}])\}^{\frac{1}{\sigma}} - 1.$$ {#eq-growth-combined}

Recall the pattern illustrated in @fig-actual-predict Panel (B): countries with higher predicted consumption growth rates tend to exhibit larger gaps between predicted and actual growth. Under the above model, the interpretation is that the combined frictions from both channels are higher in lower-income countries with high returns on both types of capital so that households rationally under-invest because the expected capital returns are lower than the realised returns.

## Accounting the Public Capital Investment Frictions {#sec-stylised}

The theoretical static model features four structural parameters: $$\Theta = \{\delta, \sigma, \tau, \phi\},$$ where $\delta$ is the annual discount factor, $\sigma$ is the risk aversion coefficient and $\frac{1}{\sigma}$ is the elasticity of intertemporal substitution, $\tau$ is the share of total investment in public capital, and $\phi$ is the severity of the collective action problem for public capital investment. To explore the data under the static model, I discipline the model by exogenously fix $\delta = 0.96$ and $\sigma = 2$ proposed by @kremer2019 based on the literature, same with the exercise in @fig-actual-predict. Although in the theoretical model, $\tau$ is chosen by the households, the public capital investment shares from ICSD can be interpretated as the equilibrium outcome of this choice.

To use macroeconomic data, I proxy for the returns on private and public capitals households face by the following Cobb-Douglas production function facing a representative household in the economy: $$ Y_{it} = A_{it}(K_{it}^{priv})^{\alpha^{priv}}(K_{it}^{pub})^{\alpha^{pub}}L_{it}^{1-\alpha^{priv}-\alpha^{pub}}, $$ where the first-order conditions are: $$ f'(K_{it}^{priv}) = \alpha_{priv} \cdot \frac{Y_{it}}{K_{it}^{priv}} \quad \text{and} \quad f'(K_{it}^{pub}) = \alpha_{pub} \cdot \frac{Y_{it}}{K_{it}^{pub}}. $$ {#eq-returns} I use the calibrate using the results from @bom2014 to fix the output elasticity of public capital while allowing the output elasticity of private capital to move mechanically. For the short-run output elasticity of private capital, assuming for simplicity the labour market is competitive in aggregate: $$w_{it} = \frac{\partial Y_{it}}{\partial L_{it}} = (1 - \alpha^{pub}_{short} - \alpha^{priv}_{it})\frac{Y_{it}}{L_{it}},$$ so that: $$\alpha^{priv}_{short,it} = 1 - \alpha^{pub}_{short} - \left(\frac{wL}{Y}\right)_{it},$$ {#eq-a-short} where the PWT provides the data on the labour share of income which is equivalent to the output elasticity of labour. Using the conservative national short-run estimate of 0.083 for $\alpha^{pub}$, this approach derives a pooled mean output elasticity of labour of 0.523, and a pooled mean ouput elasticity of private capital of 0.394. @bom2014 merge the cointegrated relationship variable and models featuring long differences variables and estimate the mean long-run elasticity of 0.122 based on this composite variable. This is consistent with the nature of longer time horizons for public projects like infrastructures and therefore longer time horizon for the returns to materialise. To capture this nature, the long-run elasticity of private capital should be calculated over every 5 year non-overlapping horizon: $$\alpha^{priv}_{long,i,[t, t+4]} = 1 - \alpha^{pub}_{long} - \left(\frac{wL}{Y}\right)_{i, [t, t+4]},$$ {#eq-a-long} and this derives a pooled mean output elasticity of labour of 0.518, and a pooled mean output elasticity of 0.36.

```{r}
#| label: read-icsd

icsd <- read_csv(here("data", "icsd2021", "dataset_2026-02-17T17_28_55.327898221Z_DEFAULT_INTEGRATION_IMF.FAD_ICSD_1.0.0.csv"))

icsd <- icsd %>%
  mutate(
    iso3c = countrycode(
      COUNTRY,
      origin = "country.name",
      destination = "iso3c",
      custom_match = c(
      "Aruba, Kingdom of the Netherlands" = "ABW",
      "Hong Kong Special Administrative Region, People's Republic of China" = "HKG",
      "Macao Special Administrative Region, People's Republic of China" = "MAC"
      )
    )
  )
```

```{r}
#| label: data-f-diff

icsd_capital <- icsd %>%
  filter(INDICATOR %in% c(
    "Capital stock, General government, Constant prices, Purchasing power parity (PPP) international dollar, ICP benchmark 2017",
    "Capital stock, Private sector, Constant prices, Purchasing power parity (PPP) international dollar, ICP benchmark 2017",
    "Capital stock, Public private partnership, Constant prices, Purchasing power parity (PPP) international dollar, ICP benchmark 2017"
  )) %>%
  mutate(
    var = case_when(
      str_detect(INDICATOR, "General government") ~ "k_pub",
      str_detect(INDICATOR, "Public private partnership") ~ "k_ppp",
      str_detect(INDICATOR, "Private sector") ~ "k_priv"
    ),
    value = as.numeric(OBS_VALUE)
  ) %>%
  select(iso3c, TIME_PERIOD, var, value) %>%
  pivot_wider(names_from = var, values_from = value)

f_diff <- pwt %>%
  filter(!is.na(labsh)) %>%
  mutate(gdp_pc = rgdpo / pop) %>%
  left_join(icsd_capital, by = c("countrycode" = "iso3c", "year" = "TIME_PERIOD")) %>%
  filter(!is.na(k_pub), !is.na(k_priv), k_pub > 0, k_priv > 0) %>%
  
  # Create 5 year horizon variables
  group_by(countrycode) %>%
  arrange(year) %>%
  mutate(
    window = floor((year - min(year)) / 5),
    n_in_window = n()
  ) %>%
  group_by(countrycode, window) %>%
  mutate(
    labsh_5y = mean(labsh, na.rm = TRUE),
    rgdpo_5y = mean(rgdpo, na.rm = TRUE),
    k_pub_5y = mean(k_pub, na.rm = TRUE),
    k_priv_5y = mean(k_priv, na.rm = TRUE),
    n_in_window = n()
  ) %>%
  ungroup() %>%
  
  mutate(
    k_ppp = replace_na(k_ppp, 0),
    
    # Short-run
    alpha_pub = 0.083,
    alpha_priv = 1 - alpha_pub - labsh,
    phi = (alpha_priv / alpha_pub) * (k_pub / k_priv),
    f_pub = alpha_pub * (rgdpo / (k_pub * 1000)),
    f_priv = alpha_priv * (rgdpo / (k_priv * 1000)),
    
    # Long-run
    alpha_pub_long = 0.122,
    alpha_priv_long = 1 - alpha_pub_long - labsh_5y,
    phi_long = (alpha_priv_long / alpha_pub_long) * (k_pub_5y / k_priv_5y),
    f_pub_long = alpha_pub_long * (rgdpo_5y / (k_pub_5y * 1000)),
    f_priv_long = alpha_priv_long * (rgdpo_5y / (k_priv_5y * 1000))
  )
```

### Difficulty to Isolate the Public Channel {#sec-selfemp}

Ideally, $\phi^{pub}$ may be approximated by a variable that is associated with the public capital investment channel but not with the private channel, and @sec-capital-public and @besley2013 show that it should likely relate to government taxation capacities. The self-employment share in total employment is a natural candidate in this context. @jensen2022 identifies that the income tax exemption threshold co-moves with the employee/self-employed composition over development, such that the tax base consistently covers segments where third-party enforceable employee income predominates (85–95% employee share). This pattern is consistent with @tbl-selfemp Panel (A) which shows higher self-employment percentage in total employment significantly associates with lower log public capital investment across specifications. However, @tbl-selfemp Panel (B) shows a highly similar pattern for the private capital investment channel. This means either self-employment share is correlated with both the private and public capital investment returns through channels like financial institution, or the frictions from both sides are highly correlated. Notice that under @eq-ff, correlation between the private and public capital returns implies correlation between the frictions from both sides. In either case, self-employment share is likely not a good proxy to distinguish $\phi^{pub}$ from the overall investment friction. Identifying isolated effects from specific channels is commonly challenging using aggregate data. Therefore attempting alternative identification strategies is beyond the scope of this paper. Nonetheless, this empirical finding motivates a follow-up test on the correlation between both friction channels.

```{r}
#| label: data-self-employment
#| eval: true

self_emp <- read_csv(
  here("data", "wdi", "API_SL", "API_SL.EMP.SELF.ZS_DS2_en_csv_v2_1885.csv"),
  skip = 3, show_col_types = FALSE
  ) %>%
  select(`Country Code`, `1991`:`2023`) %>%
  pivot_longer(
    cols = -`Country Code`,
    names_to = "year",
    values_to = "self_emp"
  ) %>%
  mutate(year = as.numeric(year)) %>%
  rename(iso3c = `Country Code`)

icsd <- icsd %>%
  left_join(self_emp, by = c("iso3c" = "iso3c", "TIME_PERIOD" = "year"))

invest_selfemp <- icsd %>%
  filter(INDICATOR %in% c(
    "Gross fixed capital formation, General government, Constant prices, Purchasing power parity (PPP) international dollar, ICP benchmark 2017",
    "Gross fixed capital formation, Private sector, Constant prices, Purchasing power parity (PPP) international dollar, ICP benchmark 2017"
  )) %>%
  
  mutate(
    var = ifelse(str_detect(INDICATOR, "General government"), "igov", "ipriv"),
    year = as.numeric(TIME_PERIOD),
    value = as.numeric(OBS_VALUE)
  ) %>%
  
  select(iso3c, year, var, value, self_emp) %>%
  pivot_wider(names_from = var, values_from = value) %>%
  left_join(
    pwt %>% select(countrycode, year, pop),
    by = c("iso3c" = "countrycode", "year" = "year")
  ) %>%
  
  group_by(iso3c) %>%
  mutate(
    igov_pc = igov / pop,
    ipriv_pc = ipriv / pop
  )
```

```{r}
#| label: tbl-selfemp
#| tbl-cap: "Self-Employment and Capital Investment"

reg_data <- invest_selfemp %>%
  ungroup() %>%
  filter(!is.na(self_emp), !is.na(igov_pc), !is.na(ipriv_pc),
         igov_pc > 0, ipriv_pc > 0)

se_models <- list(
  "Pooled" = feols(log(igov_pc) ~ self_emp, vcov = ~iso3c, data = reg_data),
  "Country FE" = feols(log(igov_pc) ~ self_emp | iso3c, vcov = ~iso3c, data = reg_data),
  "Year FE" = feols(log(igov_pc) ~ self_emp | year, vcov = ~iso3c, data = reg_data),
  "Country + Year FE" = feols(log(igov_pc) ~ self_emp | iso3c + year, vcov = ~iso3c, data = reg_data)
)

se_models_priv <- list(
  "Pooled" = feols(log(ipriv_pc) ~ self_emp, vcov = ~iso3c, data = reg_data),
  "Country FE" = feols(log(ipriv_pc) ~ self_emp | iso3c, vcov = ~iso3c, data = reg_data),
  "Year FE" = feols(log(ipriv_pc) ~ self_emp | year, vcov = ~iso3c, data = reg_data),
  "Country + Year FE" = feols(log(ipriv_pc) ~ self_emp | iso3c + year, vcov = ~iso3c, data = reg_data)
)

coef_names_se <- c("self_emp" = "Self-Employment (\\%)")
stars <- c('*' = .05, '**' = .01, '***' = .001)
gof <- data.frame(raw = c("nobs", "r2.within"), clean = c("Sample Size", "Within $R^2$"), fmt = c("%.0f", "%.3f"))

extract_tbl <- function(mods) {
  df <- modelsummary(mods, coef_map = coef_names_se, stars = stars, gof_map = gof, output = "data.frame")
  df$term[df$statistic == "std.error"] <- ""
  df %>% select(-part, -statistic)
}

tbl_a <- extract_tbl(se_models)
tbl_b <- extract_tbl(se_models_priv)

tbl_a$term <- ifelse(tbl_a$term != "", paste0("\\quad ", tbl_a$term), "")
tbl_b$term <- ifelse(tbl_b$term != "", paste0("\\quad ", tbl_b$term), "")

hdr <- \(s) tibble(term = s, `Pooled` = "", `Country FE` = "", `Year FE` = "", `Country + Year FE` = "")
out <- bind_rows(hdr("\\textbf{Panel A:} $\\log I^{pub}_{it}$"), tbl_a,
                 hdr("\\textbf{Panel B:} $\\log I^{priv}_{it}$"), tbl_b)

n_a <- nrow(tbl_a)
n_b <- nrow(tbl_b)
panel_a_row <- 1
panel_b_row <- 1 + n_a + 1

out %>%
  kbl(booktabs = TRUE, escape = FALSE, align = "lcccc",
      col.names = c("", "Pooled", "Country FE", "Year FE", "Country + Year FE")) %>%
  kable_styling(latex_options = "HOLD_position") %>%
  row_spec(panel_a_row, hline_after = TRUE) %>%
  row_spec(panel_a_row + n_a, hline_after = TRUE) %>%
  row_spec(panel_b_row, hline_after = TRUE) %>%
  row_spec(panel_b_row + n_b, hline_after = TRUE) %>%
  gsub("\\\\end\\{tabular\\}",
       "\\\\end{tabular}\n\\\\begin{minipage}{\\\\textwidth}\n\\\\vspace{2pt}\n\\\\footnotesize \\\\textit{Notes:} Dependent variable is log per capita investment. Standard errors clustered at the country level in parentheses. $^{*}p<0.05$, $^{**}p<0.01$, $^{***}p<0.001$.\n\\\\end{minipage}", .)
```

### Limited Cross-Channel Frictions Spillovers

Countries with high self-employment shares tend to be lower-income countries that share characteristics like prevalent liquidity constraints, property rights insecurity, or institutional ineffectiveness that could affect both the private and public capital investment frictions simultaneously. If this were the case, it would be a mistake to treate the two channels separately in any empirical work. Computing the marginal capital returns according to @eq-returns and test for dynamic relationship between them is informative of the correlation between the frictions from each side under @eq-ff. This is because the capital returns change due to investment behaviours correlated to investment frictions from the two channels, therefore dynamic relationship between private and public capital returns can signal correlation between the friction channels. However, one caveat is that dynamic relations can also occur due to the upward bias from common trends that do not necessarily create associations between the friction channels, but this also means that the real association is likely below the estimated levels.

@tbl-granger shows Granger tests on first-differenced \footnote{Im-Pesaran-Shin panel tests fail to reject the null of a unit root for both log $f'(K^{priv})$ ($p = 0.28$) and log $f'(K^{pub})$ ($p \approx 1.00$) in levels, but strongly reject in first differences ($p < 0.001$ for both).} log short-run marginal returns from private and public capital. Changes in public capital marginal returns exhibit significant positive autocorrelation (lagged own coefficient 0.17–0.23), while the private side does not. This asymmetry suggests slower adjustment in public investment, meaning that public capital investment frictions are sticker, while the market adjusts faster on the private side. This is consistent with factors like political inertia, long infrastructure project duration, discontinuous government budgets which do not impact private-side investments. There are statistically significant small spillovers from the public capital investment frictions to the private side, but the reverse are even smaller in magnitude and less robust across specifications. The tiny within $R^2$ under country and year fixed effects also indicate that lagged cross-channel and own-channel returns explain a small fraction of within-country variation in marginal returns at annual frequency.

@tbl-granger-long shows the Granger tests results for long-run private and public capital returns over five-year windows. The coefficients on own-channel returns are larger and more significant compared to the short run across all specifications. The stickier own-channel persistence in the long run is likely due to the five-year window reflecting more long-term structural economic trends like industrial, institutional, or TFP trajectories. However, although the private-to-public spillovers are still small, the magnitude almost doubles and the shrinking significance levels are likely resulted from shrinking sample size. This is likely also related to the coefficient absorbing the long-term trends. The long-run public-to-private spillovers attenuate instead. The increasing within $R^2$ under country and year fixed effect indicate that the long-run model has a better explanatory power.

Recall that the cross-channel spillovers can be over-estimated based on the current models by omitting common investment and structural economic trends. Nonetheless, the small magnitude of cross-channel spillovers likely indicate that the correlation between the two investment channels are likely small. Under the country and year fixed effects, both short-and long-run spillover coefficients are not significant when controlling for total factor productivity, while the Granger tests on log first-differenced capital accumulation level estimate the largest coefficient of public-to-private spillovers of 0.177 ($p < 0.1$) which is still small in magnitude.\footnote{See Appendix Table A2.} The overall empirical evidence suggest although spillovers between the private and public friction channels exist, the two investment channels and by extension the frictions operate largely via separate mechanisms. Therefore, studying the public and private capital investment channels separately is important based on this finding.

```{r}
#| label: tbl-granger
#| tbl-cap: "Short-Run Granger Tests"

f_diff_panel <- panel(f_diff, ~countrycode + year)

# Panel A: DV = f'(K^priv), Panel B: DV = f'(K^pub)
specs <- list(
  priv = d(log(f_priv)) ~ l(d(log(f_pub)), 1) + l(d(log(f_priv)), 1),
  pub  = d(log(f_pub))  ~ l(d(log(f_priv)), 1) + l(d(log(f_pub)), 1)
)
fes <- list("Pooled" = ~1, "Country FE" = ~countrycode, "Year FE" = ~year, "Country + Year FE" = ~countrycode + year)

run_panel <- function(fml) {
  base_rhs <- as.character(fml)[3]  # "l(log(f_pub), 1) + l(log(f_priv), 1)"
  dv <- as.character(fml)[2]
  list(
    "Pooled" = feols(as.formula(paste(dv, "~", base_rhs)), data = f_diff_panel),
    "Country FE" = feols(as.formula(paste(dv, "~", base_rhs, "| countrycode")), data = f_diff_panel),
    "Year FE" = feols(as.formula(paste(dv, "~", base_rhs, "| year")), data = f_diff_panel),
    "Country + Year FE" = feols(as.formula(paste(dv, "~", base_rhs, "| countrycode + year")), data = f_diff_panel)
  )
}

models_a <- run_panel(specs$priv)
models_b <- run_panel(specs$pub)

coef_names <- c("l(d(log(f_pub)), 1)" = "$\\Delta log f'(K^{pub}_{i,t-1})$",
                "l(d(log(f_priv)), 1)" = "$\\Delta log f'(K^{priv}_{i,t-1})$")
stars <- c('*' = .05, '**' = .01, '***' = .001)
gof <- data.frame(raw = c("nobs", "r2.within"), clean = c("Sample Size", "Within $R^2$"), fmt = c("%.0f", "%.3f"))

extract_tbl <- function(mods) {
  df <- modelsummary(mods, coef_map = coef_names, stars = stars, gof_map = gof, output = "data.frame")
  df$term[df$statistic == "std.error"] <- ""
  df %>% select(-part, -statistic)
}

tbl_a <- extract_tbl(models_a)
tbl_b <- extract_tbl(models_b)

# Indent coefficient/se rows
tbl_a$term <- ifelse(tbl_a$term != "", paste0("\\quad ", tbl_a$term), "")
tbl_b$term <- ifelse(tbl_b$term != "", paste0("\\quad ", tbl_b$term), "")

# Panel headers
hdr <- \(s) tibble(term = s, `Pooled` = "", `Country FE` = "", `Year FE` = "", `Country + Year FE` = "")
out <- bind_rows(hdr("\\textbf{Panel A:} $\\Delta log f'(K^{priv}_{it})$"), tbl_a,
                 hdr("\\textbf{Panel B:} $\\Delta logf'(K^{pub}_{it})$"), tbl_b)

# Row indices (1-indexed within kbl body)
n_a <- nrow(tbl_a)
n_b <- nrow(tbl_b)
panel_a_row <- 1
panel_b_row <- 1 + n_a + 1

out %>%
  kbl(booktabs = TRUE, escape = FALSE, align = "lcccc",
      col.names = c("", "Pooled", "Country FE", "Year FE", "Country + Year FE")) %>%
  kable_styling(latex_options = "hold_position") %>%
  row_spec(panel_a_row, hline_after = TRUE) %>%
  row_spec(panel_a_row + n_a, hline_after = TRUE) %>%
  row_spec(panel_b_row, hline_after = TRUE) %>%
  row_spec(panel_b_row + n_b, hline_after = TRUE) %>%
  row_spec(panel_b_row + n_b, hline_after = TRUE) %>%
  gsub("\\\\end\\{tabular\\}", 
       "\\\\end{tabular}\n\\\\begin{minipage}{\\\\textwidth}\n\\\\vspace{2pt}\n\\\\footnotesize \\\\textit{Notes:} Short-run specification uses Bom-Ligthart (2014) short-run elasticity 0.083. First difference removes AR(1) structure. Standard errors clustered at the country level in parentheses. $^{*}p<0.05$, $^{**}p<0.01$, $^{***}p<0.001$.\n\\\\end{minipage}", .)
```

```{r}
#| label: tbl-granger-long
#| tbl-cap: "Long-Run Granger Tests"

f_diff_collapsed <- f_diff %>%
  group_by(countrycode, window) %>%
  dplyr::summarise(
    f_pub_long = first(f_pub_long),
    f_priv_long = first(f_priv_long),
    .groups = "drop"
  )

f_diff_long_panel <- panel(f_diff_collapsed, ~countrycode + window)

# Panel A: DV = f'(K^priv) long, Panel B: DV = f'(K^pub) long
specs_long <- list(
  priv = d(log(f_priv_long)) ~ l(d(log(f_pub_long)), 1) + l(d(log(f_priv_long)), 1),
  pub  = d(log(f_pub_long))  ~ l(d(log(f_priv_long)), 1) + l(d(log(f_pub_long)), 1)
)

run_panel_long <- function(fml) {
  base_rhs <- as.character(fml)[3]
  dv <- as.character(fml)[2]
  list(
    "Pooled" = feols(as.formula(paste(dv, "~", base_rhs)), data = f_diff_long_panel),
    "Country FE" = feols(as.formula(paste(dv, "~", base_rhs, "| countrycode")), data = f_diff_long_panel),
    "Year FE" = feols(as.formula(paste(dv, "~", base_rhs, "| window")), data = f_diff_long_panel),
    "Country + Year FE" = feols(as.formula(paste(dv, "~", base_rhs, "| countrycode + window")), data = f_diff_long_panel)
  )
}

models_a_long <- run_panel_long(specs_long$priv)
models_b_long <- run_panel_long(specs_long$pub)

coef_names_long <- c("l(d(log(f_pub_long)), 1)" = "$\\Delta log f'(K^{pub}_{i, [t-5, t-1]})$",
                     "l(d(log(f_priv_long)), 1)" = "$\\Delta log f'(K^{priv}_{i,[t-5, t-1]})$")
stars <- c('*' = .05, '**' = .01, '***' = .001)
gof <- data.frame(raw = c("nobs", "r2.within"), clean = c("Sample Size", "Within $R^2$"), fmt = c("%.0f", "%.3f"))

extract_tbl <- function(mods, coef_map) {
  df <- modelsummary(mods, coef_map = coef_map, stars = stars, gof_map = gof, output = "data.frame")
  df$term[df$statistic == "std.error"] <- ""
  df %>% select(-part, -statistic)
}

tbl_a_long <- extract_tbl(models_a_long, coef_names_long)
tbl_b_long <- extract_tbl(models_b_long, coef_names_long)

tbl_a_long$term <- ifelse(tbl_a_long$term != "", paste0("\\quad ", tbl_a_long$term), "")
tbl_b_long$term <- ifelse(tbl_b_long$term != "", paste0("\\quad ", tbl_b_long$term), "")

hdr <- \(s) tibble(term = s, `Pooled` = "", `Country FE` = "", `Year FE` = "", `Country + Year FE` = "")
out_long <- bind_rows(hdr("\\textbf{Panel A:} $\\Delta log f'(K^{priv}_{it})$"), tbl_a_long,
                      hdr("\\textbf{Panel B:} $\\Delta log f'(K^{pub}_{it})$"), tbl_b_long)

n_a <- nrow(tbl_a_long)
n_b <- nrow(tbl_b_long)
panel_a_row <- 1
panel_b_row <- 1 + n_a + 1

out_long %>%
  kbl(booktabs = TRUE, escape = FALSE, align = "lcccc",
      col.names = c("", "Pooled", "Country FE", "Year FE", "Country + Year FE")) %>%
  kable_styling(latex_options = "hold_position") %>%
  row_spec(panel_a_row, hline_after = TRUE) %>%
  row_spec(panel_a_row + n_a, hline_after = TRUE) %>%
  row_spec(panel_b_row, hline_after = TRUE) %>%
  row_spec(panel_b_row + n_b, hline_after = TRUE) %>%
  gsub("\\\\end\\{tabular\\}", 
       "\\\\end{tabular}\n\\\\begin{minipage}{\\\\textwidth}\n\\\\vspace{2pt}\n\\\\footnotesize \\\\textit{Notes:} Long-run specification uses 5-year averaged variables and Bom-Ligthart (2014) national long-run elasticity 0.122. Year FE is applied according to the 5-year windows. First difference removes AR(1) structure. Standard errors clustered at the country level in parentheses. $^{*}p<0.05$, $^{**}p<0.01$, $^{***}p<0.001$.\n\\\\end{minipage}", .)
```

### Relative Sizes of Frictions Along Development {#sec-diff-pp}

It is difficult to decompose consumption growth induced by private and public capital investment. However, recall the equilibrium condition of @eq-ff, I can rewrite: $$\frac{\phi_{it}^{pub}}{\phi_{it}^{priv}} = \frac{f'(K_{it}^{priv})}{f'(K_{it}^{pub})} = \frac{\alpha^{priv}}{\alpha^{pub}_{it}} \left(\frac{K_{it}^{pub}}{K_{it}^{priv}}\right)$$ Using the above condition and the national and short-run output elasticity of public capital 0.083 in @bom2014, @fig-f-diff Panel (A) shows most countries across time have larger public-side wedge compared to the private side, where the marginal returns on private capital investment is higher than public capital investment. The interpretation would be that the private capital investment frictions might be higher on average across levels of development. However, the red OLS fit shows that this pattern is more prominent on the lower log per capital GDP end, and it attenuates towards the higher end.

Using instead the five-year specification and long-run output elasticity of public capital of 0.122, @fig-f-diff Panel (B) shows more even ratio between private and public side wedge across levels of development compared to Panel (A). The OLS fit exhibits a similar pattern but the public-side wedge becomes substantially smaller than the private-side wedge in higher income countries.

The important caveat for this exercise is that

Recall @fig-actual-predict that lower-income countries tend to have higher overall frictions, @fig-f-diff may suggest that the public capital investment frictions contribute less to the overall high frictions in lower-income countries, and the private-side frictions are more likely to be the driving force.

```{r}
#| label: fig-f-diff
#| fig-cap: "Ratio of Private to Public Capital Returns by Log GDP per Capita"
#| fig-cap-location: top
#| fig-height: 8

# Panel A: Short-run
diff_a <- f_diff %>%
  filter(!is.na(phi), !is.na(gdp_pc), phi > 0) %>%
  ggplot(aes(x = log(gdp_pc), y = log(phi))) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "blue") +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(
    title = "(A) Short-Run",
    x = "Log GDP per Capita",
    y = "Log Ratio of Private to Public Capital Returns"
  ) +
  theme_classic() +
  theme(
    plot.margin = margin(t = 5, r = 15, b = 5, l = 5),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5),
    axis.ticks.length = unit(-0.15, "cm"),
    axis.text.x = element_text(margin = margin(t = 0.3, unit = "cm")),
    axis.text.y = element_text(margin = margin(r = 0.3, unit = "cm")),
    plot.title = element_text(hjust = 0.5, size = 11, face = "plain")
  )

# Panel B: Long-run (5-year averages)
diff_b <- f_diff %>%
  filter(!is.na(phi_long), !is.na(gdp_pc), phi_long > 0) %>%
  ggplot(aes(x = log(gdp_pc), y = log(phi_long))) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "blue") +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(
    title = "(B) Long-Run (5-Year Averages)",
    x = "Log GDP per Capita",
    y = "Log Ratio of Private to Public Capital Returns",
    caption = stringr::str_wrap("Notes: Panel (A) uses contemporaneous values with short-run output elasticity of public capital calibrated at 0.083. Panel (B) uses 5-year averages with long-run output elasticity of public capital calibrated at 0.122. Dashed line indicates ratio of 1 (log = 0). Red line is OLS fit. Sources: PWT 11.0, ICSD.", width = 100)
  ) +
  theme_classic() +
  theme(
    plot.margin = margin(t = 5, r = 15, b = 5, l = 5),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5),
    axis.ticks.length = unit(-0.15, "cm"),
    axis.text.x = element_text(margin = margin(t = 0.3, unit = "cm")),
    axis.text.y = element_text(margin = margin(r = 0.3, unit = "cm")),
    plot.caption = element_text(hjust = 0, size = 8, face = "italic"),
    plot.caption.position = "plot",
    plot.title = element_text(hjust = 0.5, size = 11, face = "plain")
  )

diff_a / diff_b
```

## Implications

\clearpage

## References {.unnumbered}

::: {#refs}
:::

\clearpage

\appendix

\renewcommand{\thetable}{A\arabic{table}}

\setcounter{table}{0}

\renewcommand{\thefigure}{A\arabic{figure}}

\setcounter{figure}{0}

## Appendix {.unnumbered}

```{r}
#| label: tbl-poly-fit-controls
#| tbl-cap: "Polynomial Fit of Actual on Predicted Consumption Growth, with Controls"
#| tbl-pos: "H"

reg_data <- pwt_dev %>%
  group_by(countrycode) %>%
  arrange(year) %>%
  mutate(
    log_y_pc = log(gdp_pc),
    log_y_growth = log_y_pc - dplyr::lag(log_y_pc),
    log_tfp_growth = log(rwtfpna) - dplyr::lag(log(rwtfpna))
  ) %>%
  ungroup() %>%
  filter(!is.na(log_y_growth), !is.na(log_tfp_growth),
         !is.na(log_g_predicted), !is.na(log_g_actual))

poly_models <- list(
  "Linear" = lm(log_g_actual ~ log_g_predicted + log_y_growth + log_tfp_growth, 
                data = reg_data),
  "Quadratic" = lm(log_g_actual ~ poly(log_g_predicted, 2) + log_y_growth + log_tfp_growth, 
                   data = reg_data),
  "Cubic" = lm(log_g_actual ~ poly(log_g_predicted, 3) + log_y_growth + log_tfp_growth, 
               data = reg_data)
)

coef_names <- c(
  "log_g_predicted" = "Log Predicted Growth",
  "poly(log_g_predicted, 2)1" = "Poly term 1",
  "poly(log_g_predicted, 2)2" = "Poly term 2",
  "poly(log_g_predicted, 3)1" = "Poly term 1",
  "poly(log_g_predicted, 3)2" = "Poly term 2",
  "poly(log_g_predicted, 3)3" = "Poly term 3",
  "log_y_growth" = "Log GDP Growth",
  "log_tfp_growth" = "Log TFP Growth",
  "(Intercept)" = "Intercept"
)

stars <- c('*' = .05, '**' = .01, '***' = .001)
gof <- data.frame(
  raw = c("nobs", "r.squared"),
  clean = c("Sample Size", "$R^2$"),
  fmt = c("%.0f", "%.4f")
)

modelsummary(
  poly_models,
  coef_map = coef_names,
  stars = stars,
  gof_map = gof,
  output = "kableExtra",
  escape = FALSE,
  align = "lccc"
) %>%
  add_header_above(c(" " = 1, "Log Actual Consumption Growth" = 3)) %>%
  kable_styling(latex_options = "HOLD_position") %>%
  gsub("\\\\end\\{tabular\\}",
       "\\\\end{tabular}\n\\\\begin{minipage}{\\\\textwidth}\n\\\\vspace{2pt}\n\\\\footnotesize \\\\textit{Notes:} Orthogonal polynomial regressions of log actual on log predicted consumption growth, controlling for contemporaneous log GDP growth and log TFP growth. Standard errors in parentheses. \n\\\\end{minipage}", .)
```

```{r}
#| label: data-granger-robust
#| tbl-cap: "Cross-Channel Granger Tests Across Specifications"

f_diff <- f_diff %>%
  group_by(countrycode, window) %>%
  mutate(
    rwtfpna_5y = mean(rwtfpna, na.rm = TRUE),
    k_pub_5y_mean = mean(k_pub, na.rm = TRUE),
    k_priv_5y_mean = mean(k_priv, na.rm = TRUE)
  ) %>%
  ungroup()

# Short-run panel
panel_short <- panel(
  f_diff %>% filter(!is.na(rwtfpna), rwtfpna > 0),
  ~countrycode + year
)

# Long-run returns panel
f_diff_long_collapsed <- f_diff %>%
  filter(!is.na(rwtfpna_5y), rwtfpna_5y > 0,
         f_pub_long > 0, f_priv_long > 0) %>%
  group_by(countrycode, window) %>%
  dplyr::summarise(
    f_pub_long = first(f_pub_long),
    f_priv_long = first(f_priv_long),
    rwtfpna_5y = first(rwtfpna_5y),
    .groups = "drop"
  )

panel_long_returns <- panel(f_diff_long_collapsed, ~countrycode + window)

# Long-run K panel 
f_diff_long_K <- f_diff %>%
  filter(!is.na(k_pub_5y_mean), !is.na(k_priv_5y_mean),
         k_pub_5y_mean > 0, k_priv_5y_mean > 0) %>%
  group_by(countrycode, window) %>%
  dplyr::summarise(
    k_pub_5y = first(k_pub_5y_mean),
    k_priv_5y = first(k_priv_5y_mean),
    .groups = "drop"
  )

panel_long_K <- panel(f_diff_long_K, ~countrycode + window)

# Run all 8 models
m_priv_short_tfp <- feols(
  d(log(f_priv)) ~ l(d(log(f_pub)), 1) + l(d(log(f_priv)), 1) + d(log(rwtfpna)) | countrycode + year,
  data = panel_short
)

m_pub_short_tfp <- feols(
  d(log(f_pub)) ~ l(d(log(f_priv)), 1) + l(d(log(f_pub)), 1) + d(log(rwtfpna)) | countrycode + year,
  data = panel_short
)

m_priv_long_tfp <- feols(
  d(log(f_priv_long)) ~ l(d(log(f_pub_long)), 1) + l(d(log(f_priv_long)), 1) + d(log(rwtfpna_5y)) | countrycode + window,
  data = panel_long_returns
)

m_pub_long_tfp <- feols(
  d(log(f_pub_long)) ~ l(d(log(f_priv_long)), 1) + l(d(log(f_pub_long)), 1) + d(log(rwtfpna_5y)) | countrycode + window,
  data = panel_long_returns
)

m_priv_K <- feols(
  d(log(k_priv)) ~ l(d(log(k_pub)), 1) + l(d(log(k_priv)), 1) | countrycode + year,
  data = panel_short
)

m_pub_K <- feols(
  d(log(k_pub)) ~ l(d(log(k_priv)), 1) + l(d(log(k_pub)), 1) | countrycode + year,
  data = panel_short
)

m_priv_K_long <- feols(
  d(log(k_priv_5y)) ~ l(d(log(k_pub_5y)), 1) + l(d(log(k_priv_5y)), 1) | countrycode + window,
  data = panel_long_K
)

m_pub_K_long <- feols(
  d(log(k_pub_5y)) ~ l(d(log(k_priv_5y)), 1) + l(d(log(k_pub_5y)), 1) | countrycode + window,
  data = panel_long_K
)

models_list <- list(
  "$\\Delta\\log f'(K^{priv})$ Short" = m_priv_short_tfp,
  "$\\Delta\\log f'(K^{pub})$ Short" = m_pub_short_tfp,
  "$\\Delta\\log f'(K^{priv})$ Long" = m_priv_long_tfp,
  "$\\Delta\\log f'(K^{pub})$ Long" = m_pub_long_tfp,
  "$\\Delta\\log K^{priv}$ Short" = m_priv_K,
  "$\\Delta\\log K^{pub}$ Short" = m_pub_K,
  "$\\Delta\\log K^{priv}$ Long" = m_priv_K_long,
  "$\\Delta\\log K^{pub}$ Long" = m_pub_K_long
)

coef_map <- c(
  "l(d(log(f_pub)), 1)" = "$\\Delta\\log f'(K^{pub}_{i,t-1})$",
  "l(d(log(f_priv)), 1)" = "$\\Delta\\log f'(K^{priv}_{i,t-1})$",
  "l(d(log(f_pub_long)), 1)" = "$\\Delta\\log f'(K^{pub}_{i,w-1})$",
  "l(d(log(f_priv_long)), 1)" = "$\\Delta\\log f'(K^{priv}_{i,w-1})$",
  "l(d(log(k_pub)), 1)" = "$\\Delta\\log K^{pub}_{i,t-1}$",
  "l(d(log(k_priv)), 1)" = "$\\Delta\\log K^{priv}_{i,t-1}$",
  "l(d(log(k_pub_5y)), 1)" = "$\\Delta\\log K^{pub}_{i,w-1}$",
  "l(d(log(k_priv_5y)), 1)" = "$\\Delta\\log K^{priv}_{i,w-1}$",
  "d(log(rwtfpna))" = "$\\Delta\\log \\text{TFP}_{it}$",
  "d(log(rwtfpna_5y))" = "$\\Delta\\log \\text{TFP}_{iw}$"
)

stars <- c('*' = .05, '**' = .01, '***' = .001)
gof <- data.frame(
  raw = c("nobs", "r2.within"),
  clean = c("Sample Size", "Within $R^2$"),
  fmt = c("%.0f", "%.3f")
)

modelsummary(
  models_list,
  coef_map = coef_map,
  stars = stars,
  gof_map = gof,
  escape = FALSE,
  output = "kableExtra",
  notes = "Standard errors clustered at country level in parentheses. All specifications include country and year (window for long-run) fixed effects. Returns specifications include contemporaneous TFP control."
) %>%
  kable_styling(latex_options = "scale_down") %>%
  landscape()
```
